---
title: "NCAA Tournament Contest"
description: |
  Scraping Data
output:
  radix::radix_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# A Brief Review

Last time, we scraped data coming from a few different sources: tables, things that look like tables, and numbers clearly coming from Dante's 5th circle. The bulk of our work was done using the common `read_html` to `html_table` or `html_nodes`, we did see that we need to utilize a little bit of XML for some of the more complicated affairs.  

While our focus last time was mainly getting the data in, we are going to focus our attention now on cleaning the data and hoping to join it together.

# Basic Data Manipulation

```{r}
library(dplyr)

library(stringr)

library(tidyr)

load("basketballData.RData")
```


If we take another look at our data from Ken Pomeroy, we see that we might have an interesting issue.

```{r}
rmarkdown::paged_table(kenPomData)
```

We need to do something about those columns names:

```{r}
oldNames = names(kenPomData)

newNames = kenPomData[1, ]

newNames = unlist(ifelse(oldNames != "", paste(newNames, oldNames, sep = "_"), newNames))

newNames = gsub("\\s|-", "_", newNames)

newNames = ifelse(duplicated(newNames), paste(newNames, "rank", sep = "_"), newNames)

names(kenPomData) = newNames

kenPomData = kenPomData %>% 
  filter(Rk != "" & Rk != "Rk")

rmarkdown::paged_table(kenPomData)
```


Finally, we can clean up the variables and convert them appropriately:

```{r}
kenPomData = kenPomData %>% 
  separate(., W_L, into = c("Wins", "Losses")) %>%
  mutate_at(., .vars = vars(Rk, Wins:AdjEM_NCSOS_rank), as.numeric) %>% 
  mutate(Win_Percentage = Wins / (Wins + Losses))

rmarkdown::paged_table(kenPomData)
```


# String Processing

Let's begin by returning to one of our first tables that we pulled in from the web: the nittyGritty table

```{r}
head(nittyGrittyData)
```

For the sake of getting the data in shape, we need to drop those rows without names:

```{r}
dropCols = which(names(nittyGrittyData) == "")

nittyGrittyData = nittyGrittyData[, -c(dropCols)]
```

While things looked good with a cursory glance, we see some issues straight away. The major issue is that we have strings smashed together. Fortunately, this is not too big of a problem:

```{r}
pattern = "([a-z])([A-Z])"

team %>% 
  str_subset(pattern) %>% 
  str_match(pattern)

nittyGrittyData %>% 
  mutate(team = , 
         record = str_extract(Team, "[0-9]+-[0-9]+"), 
         conference)

nittyGrittyData$Team = gsub("([a-z])([A-Z])", "\\1 \\2", nittyGrittyData$Team)

nittyGrittyData$Team = gsub("\\(|\\)", "", nittyGrittyData$Team)

nittyGrittyData %>% 
  filter(Team != "Team") %>% 
  separate(., col = Team, into = c("Team", "Record"), 
           sep = "\\s(?=[0-9]+-)", remove = FALSE)

rmarkdown::paged_table(nittyGrittyData)
```

That is going to get most of the names separated from the conference, but definitely not all (the abbreviation schools are going to cause us some issues here)!



We can remove extra rows with variable names and start pulling the data out of the `Team` variable.

```{r}
nittyGrittyData$Team = gsub("\\(\\)", "", nittyGrittyData$Team)
```

